### 2013-03-23 doc/sosp13/tasks.tex
new Tasks section

\section{Tasks}

The basic unit of execution in \Grappa is a {\em task}. Each task is
represented by a function pointer and its arguments. When a new task is
spawned, it is enqueued until it is ready for execution. Each hardware core
has a single operating system thread pinned to it. A {\em worker} is a
user-level execution thread (stack, \TODO{what else?}) to which tasks are
mapped onto when they are ready to execute.

Once a task is mapped to worker it stays with that worker until it completes. 
During execution, a worker yields control of the processor core whenever it
performs a long-latency operation, allowing the processor to remain busy while
waiting for the operation to complete.
In addition, a programmer can direct
scheduling explicitly via the Grappa API calls shown in
Figure~\ref{fig:scheduling}. To minimize yield overhead, the Grappa scheduler
operates entirely in user-space and does little more than store register state
of one task and load that of another. Context switch times are as low as 38ns
even when switching amongst thousands of tasks.

\Grappa keeps a large number of workers per hardware core, which is at the
core of \Grappa's latency tolerance properties.

\TODO{We should expand on the cleverness of our scheduler?}

% talk about lifetime of a task. activating. blocking. work stealing. look at
% Mark's email. Holt's chat.

\comment{
Delete the spin and move the 2us and ms thing to results if we have data to show it. -Mark

Fast context switching amongst many tasks is
crucial to Grappa for two reasons. First, only when the net cost of context
switching is dominated by the loss incurred by stalling does a latency
tolerant approach work to increase throughput on computations with intense
random access. Random access latencies to node local memory are on the order
of 100ns. Were the scheduler to add more overhead than this, it would actually
slow down rather than speed up computations that randomly access node local
data. Second, because we can switch amongst so many tasks without dramatically
increasing scheduling overhead, Grappa can tolerate system latencies many
orders of magnitude greater than that of local memory. Network ping latency
alone is approximately 2us, but our data suggests we can tolerate latencies in the millisecond
scale. Furthermore, by over-provisioning task parallelism, Grappa promises to
provide robust throughput that is relatively insensitive to variations in
latencies within and across our target systems.
}

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{ yield() } ] \hfill \\
      Yields core to scheduler, enqueuing caller to be scheduled again soon
    \item[ \texttt{ suspend() } ] \hfill \\
      Yields core to scheduler, enqueuing caller only once another task calls wake
    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
      Enqueues some other task $t$ to be scheduled again soon
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:scheduling} Grappa API: scheduling } %{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

\subsection{Expressing parallelism}

@@

\subsection{Implementation tasks}

\paragraph{Tasks and workers} Grappa tasks are 32-byte entities: a
64-bit function pointer plus three 64-bit arguments. We use three
arguments because tasks are often generated via a parallel loop
decomposition, in which each task needs three kinds of data.
\begin{description}
\item[Function pointer] addresses the routine to run. We configure the system nodes to disable address randomization, so that function pointers are valid across process images.
\item[Private argument] often a loop index.
\item[Shared argument] typically data shared amongst a group of tasks, or the number of loop iterations to be performed by this task
\item[Synchronization argument] often used to determine
  when all tasks that are part of a loop have finished; a global pointer to a synchronization object allocated at the core that spawned a group of tasks.
\end{description}
While these are the most common uses of the three task arguments, they
are treated as arbitrary 64-bit values in the runtime, and can be used
for any purpose.

Tasks are not allocated any execution resources until the scheduler
decides to run them; when this occurs, tasks are matched with {\em
  worker} threads. Each worker is simply a collection of status bits and a
stack, allocated at a particular core.

\paragraph{Context switching} Grappa context switches between tasks
non-preemptively. As with other cooperative multithreading systems, we
treat context switches as function calls, saving and restoring only the
callee-saved state as specified in the x86-64 ABI \cite{amd64:abi:2012}. This
involves saving six general-purpose 64-bit registers and the stack
pointer, as well as the 16-bit x87 floating point control word and the
SSE context/status register. Thus, the minimum amount of state a
cooperative context switch routine must save according to the ABI is 62
bytes.

\paragraph{Scheduling} Each core in the Grappa system has its own
independent scheduler. Each scheduler has three main tasks to perform:
servicing communication requests as described in
Section~\ref{sec:communication}; rescheduling tasks that have been
waiting on long-latency operations; and assigning ready tasks to
worker resources that have become idle.

Each scheduler has three queues:
\begin{description}
\item{\bf Ready worker queue} This is a FIFO queue of tasks that are
  matched with workers and are ready to execute.
\item{\bf Private task queue} This is a FIFO queue of tasks that must run on this core.
\item{\bf Public task queue} This is a LIFO queue of tasks that are
  waiting to be matched with workers. It is a local partition of a shared
  task pool.
\end{description}

Whenever a task yields or suspends, the scheduler makes a decision
about what to do next. Servicing communication requests is given priority
to ensure responsiveness, but to minimize overhead should context switches be frequent, servicing is performed only if sufficient time has elapsed.
Second, the scheduler determines if any workers with running tasks are
ready to execute; if so, one is scheduled. Finally, if there are no
workers ready to run, but there are tasks waiting to be matched with
workers, an idle worker is woken (or a new worker is spawned), matched
with a task, and scheduled.

>>>>

\section{Tasking System}

The basic unit of execution in \Grappa is a {\em task}. Each hardware core has
a single operating system thread pinned to it. When tasks are ready to
execute, they are mapped to a {\em worker}, which is a user-level thread.
\Grappa keeps a large number of workers per hardware core, which is at the
core of \Grappa's latency tolerance properties. Below we discuss the
implementation of our task management support and then describe how
applications expose parallelism to the \Grappa runtime.

\subsection{Task Support Implementation}

\paragraph{Tasks}
Each task is represented by a function pointer and its arguments, represented
as 32-byte entities: a 64-bit function pointer plus three 64-bit arguments. We
use three arguments because tasks are often generated via a parallel loop
decomposition, in which each task needs three kinds of data. Thus, tasks
include:

\noindent {\it function pointer}, or the addresses of the routine to run ---
we configure the system nodes to disable address randomization, so that
function pointers are valid across process images;

\noindent {\it private argument}, often a loop index;

\noindent {\it shared argument}, typically data shared amongst a group of
tasks, or the number of loop iterations to be performed by this task;

\noindent {\it synchronization argument}, often used to determine when all
tasks that are part of a loop have finished, in the form of q global pointer
to a synchronization object allocated at the core that spawned a group of
tasks.

While these are the most common uses of the three task arguments, they are
treated as arbitrary 64-bit values in the runtime, and can be used for any
purpose.

\paragraph{Workers}

When a new task is spawned, it is enqueued until it is ready for execution.
Tasks are not allocated any execution resources (a worker) until the scheduler
decides to run them. A worker is simply a collection of status bits and a
stack, allocated at a particular core. There are two types of workers: (1) user worker; and (2) system worker.

User workers execute application tasks. Once a task is mapped to a worker it
stays with that worker until it completes. System workers execute runtime
management activities such as communication and statistics collection. User
workers and system workers might have different scheduling priorities. System
workers execute periodically and in idle times, to guarantee timely
communication services.

\paragraph{Scheduling}

During execution, a worker yields control of its core whenever it performs a
long-latency operation, allowing the processor to remain busy while waiting
for the operation to complete. In addition, a programmer can direct scheduling
explicitly via the Grappa API calls shown in Figure~\ref{fig:scheduling}. To
minimize yield overhead, the Grappa scheduler operates entirely in user-space
and does little more than store register state of one task and load that of
another.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{ yield() } ] \hfill \\
      Yields core to scheduler, enqueuing caller to be scheduled again soon
    \item[ \texttt{ suspend() } ] \hfill \\
      Yields core to scheduler, enqueuing caller only once another task calls wake
    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
      Enqueues some other task $t$ to be scheduled again soon
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:scheduling} \Grappa scheduling API } 
    \end{minipage}
  \end{center}
\end{figure}

Each core in a \Grappa system has its own independent scheduler. Each
scheduler has three main responsibilities: servicing system workers that
process communication requests as described in
Section~\ref{sec:communication}; rescheduling workers that have been waiting
on long-latency operations; and assigning ready tasks to worker resources that
have become idle.

Each scheduler has three queues:

\noindent {\it ready worker queue}, a FIFO queue of tasks that are
  matched with workers and are ready to execute;

\noindent {\it private task queue}, a FIFO queue of tasks that must run on this core;

\noindent {\it public task queue},  a LIFO queue of tasks that are
  waiting to be matched with workers. It is a local partition of a shared
  task pool.

\TODO{don;t we have an actual queue for higher priority system workers?}

Whenever a task yields or suspends, the scheduler makes a decision about what
to do next. Servicing communication requests (system worker) is given priority
to ensure responsiveness, but to minimize overhead should context switches be
frequent, servicing is performed only if sufficient time has elapsed. Second,
the scheduler determines if any workers with running tasks are ready to
execute; if so, one is scheduled. Finally, if there are no workers ready to
run, but there are tasks waiting to be matched with workers, an idle worker is
woken (or a new worker is spawned), matched with a task, and scheduled.

\paragraph{Context switching} Grappa context switches between workers
non-preemptively. As with other cooperative multithreading systems, we
treat context switches as function calls, saving and restoring only the
callee-saved state as specified in the x86-64 ABI \cite{amd64:abi:2012}. This
involves saving six general-purpose 64-bit registers and the stack
pointer, as well as the 16-bit x87 floating point control word and the
SSE context/status register. Thus, the minimum amount of state a
cooperative context switch routine must save, according to the ABI, is
62~bytes --- importantly, a context is saved in a single 64 cache line.

Since \Grappa keeps a very large number of active workers, their context data will not fit in cache. Therefore, the scheduler has to perform very careful prefetching in order to reduce context switch cost to a minimum. We determined empirically that prefetching the fourth worker in the scheduling order is sufficient. This makes context switching effectively free of cache misses.

% When the scheduler chooses a thread to schedule it first looks to see if it must run a periodic task.  If so it choose that.  If not it looks for a ready task.  If none exists it chooses the idle thread.
% 
% If it chooses a task from the ready queue it prefetches the 4'th next thread on the read queue.  Presently 192 bytes of stack are prefetched and 64 bytes of the thread struct itself.
% 
% A periodic task is responsible for looking at the size of the ready queue.  If it is too small more not started tasks are started.
%

% The context switcher is written entirely in user space.  A context is saved in a single cache line (64 bytes). We prefetch 192 bytes from the stack (which is where the context is saved) in order to prime the stride prefetched on the core and to preload the task state that exists on the stack.

% Fast context switching amongst many tasks is crucial to Grappa for two
% reasons. First, only when the net cost of context switching is dominated by
% the loss incurred by stalling does a latency tolerant approach work to
% increase throughput on computations with intense random access. Random access
% latencies to node local memory are on the order of 100ns. Were the scheduler
% to add more overhead than this, it would actually slow down rather than speed
% up computations that randomly access node local data. Second, because we can
% switch amongst so many tasks without dramatically increasing scheduling
% overhead, Grappa can tolerate system latencies many orders of magnitude
% greater than that of local memory. Network ping latency alone is approximately
% 2us, but our data suggests we can tolerate latencies in the millisecond scale.
% Furthermore, by over-provisioning task parallelism, Grappa promises to provide
% robust throughput that is relatively insensitive to variations in latencies
% within and across our target systems.

\subsection{Expressing Parallelism}

@@
